---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

This is the markdown file to the Peer Assessment 1 assignment. 
Note that the relevant zip file is in the same Git Hub repository; the data can be downloaded here <insert link>.

## Loading and preprocessing the data
this step loads the data in R, and formats the date variable as an R date variable.

```{r}
data <- read.csv(unz("./activity.zip", "activity.csv"))
data$date <- as.Date(data$date, "%Y-%m-%d")
head(data)
```


## Calculate total and Mean total number of steps taken per day
this step totals the number of steps taken a day (at 5 minute intervals). Note that this aggregates the daily steps for a given interval. For example - the total / mean / median for interval 1225 is the corresponding total / mean / median for the given interval, across the dates.

```{r summstats}
# melt the data, as 
library(reshape)
stepsummary <- melt(data, id = "interval")
stepsummary <- cast(stepsummary, interval ~ variable, mean) #obtain the average steps per day 
head(stepsummary)
```

### The average steps at 1225 is 'r stepsumm'.

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
